Appendix 1: Quantifying Drought

### 2021-03-03 analysis/supplementary-materials/appendix_SPEI.Rmd
Changed imputed BDFFP SPEI to only use BDFFP for precipitation and to use eto from the mean of the two EMBRAPA stations.

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  include = FALSE,
  echo = FALSE,
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "",
  dpi = 300)
library(tidyverse)
library(here)
library(conflicted)
library(SPEI)
library(tsibble)
library(lubridate)
library(glue)
library(colorspace)
library(latex2exp)

@@

*Last compiled: `r Sys.Date()`*
# TODO

- Write
- Fix colors and labels and such on fig S1
- Make a figure S2 that compares SPEI across BDFFP from east to west (3 panels)
- Maybe add imputed data to this (or maybe not)

@@

```{r read-data}
# Precip Sources
xa <- read_csv(here("analysis", "data", "raw_data", "xavier_daily_0.25x0.25.csv"))
sa <- read_csv(here("analysis", "data", "raw_data", "sa_daily_1x1.csv"))
# trmm <-

# Already calculated SPEI  
gspei <-
  read_delim(here("analysis", "data", "raw_data", "global_spei_-59.75_-2.25.csv"),
             delim = ";") %>% 
  rename(date = "days since 1900-1-1", g_spei = spei) %>% 
  mutate(yearmonth = yearmonth(date), date = NULL)

bdffp <- read_csv(here("analysis", "data", "raw_data", "mon_precip_spi_imputed.csv"))
#TODO edit to pull this from github or use bdffp rainfall data package

@@

## Data Source Descriptions

### BDFFP rain gauge data

Daily rain gauge data collected from 1987--2010 at BDFFP camps. The precipitation data has a high degree of missingness due to camps being visited infrequently and because not all camps were active from 1987--2010. There we also a large number of observations that were accumulated rainfall over two or more days. We removed multi-day accumulations as it was impossible to know how rainfall was distributed over the prior days. Missing values were then imputed after averaging rainfall data by ranch. This averaging was necessary as the degree of missingness was too high for imputation to proceed for some camps. Multiple imputation was performed using the `Amelia` package and was informed by lags/leads, seasonality, and continuous daily weather data from two nearby by weather stations including precipitation, temperature, relative humidity, and wind speed. Multiple imputation produced \_\_\_ <!--# 10? --> imputed daily datasets which were then each aggregated monthly and used for calculating SPEI using potential evapotranspiration from Xavier et al. [-@xavierDailyGriddedMeteorological2016] to calculate climatic water balance <!--# why not use mean temperature from weather stations? -->. Then, the <!--# number --> resulting SPEI datasets were averaged to produce a single dataset.

@@

-   TRMM 3B43 <!--# doesn't go back far enough.  Not worth using? -->
    -   Source DOI: 10.5067/TRMM/TMPA/MONTH/7
    -   Resolution: 0.25º x 0.25º
    -   Includes: Monthly precip
    -   Doesn't include PET, so I guess I'll take that from the ground stations??
    -   Date range: 1998--2019

@@

# Xavier et al.

```{r xa-spei}
#aggregate monthly, then calculate climate water balance.
xa_mon <-
  xa %>% 
  unite(latlon, lat, lon, sep = ", ") %>% 
  mutate(yearmonth = yearmonth(date), .after = date) %>% 
  group_by(yearmonth, latlon) %>% 
  summarize(across(c(precip, eto), sum), .groups = "drop") %>% 
  as_tsibble(index = yearmonth, key = latlon)

# EMBRAPA stations

embrapa_wide <-
  embrapa_spei %>% 
  select(yearmonth, station, st_spei) %>% 
  pivot_wider(names_from= station, values_from = st_spei, names_prefix = "spei_") %>% 
  rowwise() %>% 
  mutate(st_spei = mean(c(spei_manaus, spei_rpde))) %>% 
  ungroup()
```

# TRMM

# Imputed data (averaged across for BDFFP)

```{r}
bdffp_spei <- 
  bdffp %>% group_by(date) %>% 
  summarize(bdffp_spei = mean(spei)) %>% 
  mutate(yearmonth = yearmonth(date), date = NULL)
```

comb <-
  full_join(xa_df, gspei) %>% 
  full_join(trmm_df) %>% 
  full_join(embrapa_wide %>% select(yearmonth, st_spei)) %>% 
  full_join(bdffp_spei) %>% 
  pivot_longer(-yearmonth, names_to = "dataset", values_to = "spei") %>% 
  filter(yearmonth > yearmonth(ymd("1990-01-01"))) %>% 
  mutate(year = year(yearmonth), .before = yearmonth)
```

# Table of Expected Droughts
Annotate worst droughts.\
Expected:

>>>>

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  include = FALSE,
  echo = FALSE,
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.path = "",
  dpi = 300)
library(tidyverse)
library(here)
library(conflicted)
library(SPEI)
library(tsibble)
library(lubridate)
library(glue)
library(colorspace)
library(latex2exp)
library(janitor)
library(Amelia)

@@

*Last compiled: `r Sys.Date()`* \# TODO

-   Write
-   ~~Fix colors and labels and such on fig S1~~
-   Make a figure S2 that compares SPEI across BDFFP from east to west (3 panels)
-   ~~Maybe add imputed data to this (or maybe not)~~
-   Move imputation code to this notebook. This requires finishing up the bdffp precip R package.

@@

```{r read-data}
# Precip Sources
sa <- read_csv(here("analysis", "data", "raw_data", "sa_daily_1x1.csv"))
# trmm <-

# Already calculated SPEI

# bdffp <- read_csv(here("analysis", "data", "raw_data", "mon_precip_spi_imputed.csv"))
# #TODO edit to pull this from github or use bdffp rainfall data package

@@

We compared 3-month SPEI data calculated from a number of sources for the BDFFP site including ground-based data from on-site rain gauges, two nearby weather stations, satellite products, and gridded land station based datasets.

## EMBRAPA

embrapa_wide <-
  embrapa_spei %>% 
  select(yearmonth, station, st_spei) %>% 
  pivot_wider(names_from = station,
              values_from = st_spei,
              names_prefix = "spei_") %>% 
  rowwise() %>% 
  mutate(st_spei = mean(c(spei_manaus, spei_rpde))) %>% 
  ungroup()
```

Two weather stations within \_\_\_ km of BDFFP had data available through EMBRAPA <!--# what is this -->, Rio Preto da Eva to the East (-2.65º, -59.65º) and the Manaus station to the West (-2.65º, -60.35º).
We downloaded daily precipitation and potential evapotranspiration data, aggregated it to monthly, and calculated 3 month SPEI for each station separately.
We then averaged the SPEI values for two stations at each month to produce monthly estimates for the BDFFP site.

## BDFFP rain gauge data

```{r}
# Read in data 
bdffp_daily <- read_csv("https://raw.githubusercontent.com/BrunaLab/Heliconia-Drought/master/data_cleaned/daily_precip.csv?token=AGB2K35OCIPUILNGOD6VXV3AH7XOU")

# Prep data for imputation ------------------------------------------------

# remove accumulations
bdffp2 <-
  bdffp_daily %>% 
  filter(!flag %in% c("A", "U"))

# spread and add rows for missing dates
bdffp_wide <- 
  bdffp2 %>% 
  #complete all the dates
  as_tsibble(key = site, index = date) %>% 
  fill_gaps() %>% 
  select(date, site, precip) %>% 
  as_tibble() %>% 
  pivot_wider(names_from = site, values_from = precip) %>% 
  clean_names()

# Cluster sites by ranch to reduce missingness
bdffp_wide2 <-
  bdffp_wide %>% 
  rowwise() %>% 
  mutate(colosso_clust = mean(c(colosso, florestal, cabo_frio, gaviao),
                              na.rm = TRUE),
         km_clust = mean(c(km37, km41),
                         na.rm = TRUE)) %>% 
  mutate(across(colosso_clust:km_clust, ~ifelse(is.nan(.x), NA, .x))) %>% 
  select(-cabo_frio, -colosso, -florestal, -gaviao, -km37, -km41)

# add additional representations of time for Amelia.
full_wide <- 
  bdffp_wide2 %>% 
  mutate(
    year = year(date),
    doy = yday(date),
    yearmonth = yearmonth(date),
    .before = everything()
  )
```

```{r}
# Impute missing data -----------------------------------------------------
# round(runif(1, 1, 1000))
set.seed(937)
all_cols <- colnames(select(full_wide, -year, -doy, -date))

# histograms <-
#   map(all_cols, ~{
#     ggplot(full_wide, aes(full_wide[[.x]])) + geom_histogram() + labs(title = .x)
#   })
# histograms

# variables with log-normal-ish distributions

log_cols <-
  all_cols[str_detect(all_cols, "precip") |
             all_cols %in% c("km_clust", "colosso_clust", "porto_alegre", "dimona")]

imp <- 
  amelia(
    as.data.frame(full_wide),
    p2s = 0,
    m = 10,
    ts = "doy",
    cs = "year",
    intercs = TRUE,
    polytime = 3,
    logs = log_cols,
    idvars = c("date"),
    parallel = "multicore",
    ncpus = detectCores() - 1,
    empri = .01 * nrow(full_wide) #ridge penalty because of high degree of missingness
  )
# imp$imputations[[1]]
```

```{r}
# Prep eto data for SPEI calculation
# Aggregate by month
imp_mon <-
  map(imp$imputations, ~{
    .x %>% 
      rename_with(~paste0(., ".precip"), c(dimona, porto_alegre, colosso_clust, km_clust)) %>% 
      select(date, ends_with(".precip")) %>% 
      mutate(yearmonth = tsibble::yearmonth(date)) %>% 
      group_by(yearmonth) %>% 
      summarize(across(-date, ~sum(.x, na.rm = TRUE))) %>% 
      #remove first and last month, as they aren't complete
      slice(-1, -nrow(.))
  })
# Add mean ETO from EMBRAPA data
imp_mon <-
  map(imp_mon, ~ {
    left_join(.x, embrapa_mon %>% summarize(eto = mean(eto)), by = "yearmonth")
  }) %>% 
  map(., ~{
    mutate(.x, across(ends_with(".precip"),
                      ~ .x - eto,
                      .names = "{str_remove(col, '.precip')}.cb"))
  })
# imp_mon$imp1
```

```{r}
# Calculate SPEI3 (i.e. 3-month window SPEI)

imp_spei <-
  map(imp_mon, ~{
  .x %>% 
    mutate(across(ends_with(".cb"), ~as.numeric(SPEI::spei(.x, scale = 3)$fitted),
                  .names = "{str_remove(col, '.cb')}.spei")) %>%
      select(yearmonth, ends_with(c(".precip", ".spi", ".spei")))
}) 
# imp_spei$imp1
# Combine multiply imputed SPI and SPEI results by taking mean
bdffp_spei <- 
  imp_spei %>%
  bind_rows(.id = "imp") %>% 
  #replace any infinite values with NAs
  mutate(across(ends_with(".spei"), ~ ifelse(is.infinite(.x), NA, .x))) %>% 
  group_by(yearmonth) %>% 
  summarize(across(where(is.numeric), ~mean(., na.rm = TRUE))) %>% 
  #replace NaN's with NAs
  mutate(across(ends_with(".spei"), ~ ifelse(is.nan(.x), NA, .x))) %>% 
  rowwise() %>% 
  mutate(bdffp_spei = mean(c_across(ends_with(".spei")))) %>% 
  ungroup()
```

We obtained rain gauge data collected from 1987--2010 at BDFFP camps.
The precipitation data has a high degree of missingness due to camps being visited infrequently and because not all camps were active from 1987--2010.
There we also a large number of observations representing accumulated rainfall over two or more days.
We removed multi-day accumulations as it was impossible to know how rainfall was distributed over the prior days.
Missing values were then imputed after averaging rainfall data by ranch.
This averaging was necessary as the degree of missingness was too high for imputation to proceed for some camps.
Multiple imputation was performed using the `Amelia` package and was informed by lags/leads, seasonality, and continuous daily weather data from two nearby by weather stations including precipitation, temperature, relative humidity, and wind speed [@honakerAmeliaIIProgram2011].
We produced 10 imputed daily datasets which were then each aggregated monthly and used for calculating SPEI using the averaged potential evapotranspiration from the two EMBRAPA weather stations<!--# why not use mean temperature from weather stations? -->.
Then, the 10 resulting SPEI datasets were averaged to produce a single dataset.

## Xavier et al.

```{r xa-spei}
xa <- read_csv(here("analysis", "data", "raw_data", "xavier_daily_0.25x0.25.csv"))
#aggregate monthly, then calculate climate water balance.
xa_mon <-
  xa %>% 
  unite(latlon, lat, lon, sep = ", ") %>% 
  mutate(yearmonth = yearmonth(date), .after = date) %>% 
  group_by(yearmonth, latlon) %>% 
  summarize(across(c(precip, eto), sum), .groups = "drop") %>% 
  as_tsibble(index = yearmonth, key = latlon)

Gridded data from Xavier et al. [-@xavierDailyGriddedMeteorological2016] was used to calculate SPEI as described in the main text.
For figure S1, we used an average of SPEI values across all grid cells.

## TRMM

Monthly precipitation data from TRMM was used in combination with potential evapotranspiration data from Xavier et al. [-@xavierDailyGriddedMeteorological2016].

## Global SPEI

```{r global-spei}
gspei <-
  read_delim(here("analysis", "data", "raw_data", "global_spei_-59.75_-2.25.csv"),
             delim = ";") %>% 
  rename(date = "days since 1900-1-1", g_spei = spei) %>% 
  mutate(yearmonth = yearmonth(date), date = NULL)
```

Additionally, we downloaded SPEI data from the Global SPEI dataset [@begueriaSbegueriaSPEIbaseVersion2017].
These calculations were made using precipitation data from a gridded product based on land stations and evapotranspiration estimated with the Pennman-Montieth method.

comb <-
  full_join(xa_df, gspei) %>% 
  full_join(trmm_df) %>% 
  full_join(embrapa_wide %>% select(yearmonth, st_spei)) %>% 
  full_join(bdffp_spei %>% select(yearmonth, bdffp_spei)) %>% 
  pivot_longer(-yearmonth, names_to = "dataset", values_to = "spei") %>% 
  filter(yearmonth > yearmonth(ymd("1990-01-01"))) %>% 
  mutate(year = year(yearmonth), .before = yearmonth)
```

# Table of Expected Droughts

Annotate worst droughts.\
Expected:

# Notes

@@

-   TRMM 3B43 <!--# doesn't go back far enough.  Not worth using? -->

    -   Source DOI: 10.5067/TRMM/TMPA/MONTH/7
    -   Resolution: 0.25º x 0.25º
    -   Includes: Monthly precip
    -   Doesn't include PET, so I guess I'll take that from the ground stations??
    -   Date range: 1998--2019

@@
